---
title: "Testing the `bjlm` Package"
author: "Antigravity"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# Uncomment the following line to load the package from source during development
devtools::load_all(".")
# library(bjlm)
set.seed(42)
```

## Introduction

This document contains a series of tests for the `bjlm` package using simulated data of varying sizes. The goal is to push the sampler and the API using edge cases, unusual but plausible specifications, and common scenarios. 

This test suite covers:
1. A basic cross-sectional linear model.
2. A logistic regression model with an extreme weight distribution and a single structural breakpoint.
3. A complex longitudinal negative binomial model with subject-level random intercepts and a Latent Gaussian Process (GP) confounder.

---

## Scenario 1: Basic Linear Model

We start with a very simple cross-sectional design. We simulate a continuous outcome `Y` dependent on time `tau` and a binary treatment `Trt`.

```{r scenario1}
n <- 200
dat1 <- data.frame(id = 1:n, tau = runif(n, 0, 10), Trt = rbinom(n, 1, 0.5))
# True model: Y = 2 + 0.5*tau + 1.5*Trt + error
dat1$Y <- 2 + 0.5 * dat1$tau + 1.5 * dat1$Trt + rnorm(n)

fit1 <- bjlm_model() |>
  propensity(Trt ~ 1, data = dat1) |>
  outcome(Y ~ tau, b0 = ~ 1 + Trt, b1 = ~ 1, data = dat1) |>
  compile() |>
  fit(iter = 500, warmup = 200, chains = 2, cores = 2)

# The table should show an intercept of ~2.0, b1 of ~0.5, and Trt of ~1.5
tab_bjlm(fit1)
```

---

## Scenario 2: Logistic Regression with Breakpoint & Extreme Weights

Here we test the robustness of the Pólya-Gamma sampler and the Inverse Probability Weight (IPW) truncation. We generate an extreme propensity score distribution, leading to large weights, and limit `max_weight` to 10. We also simulate a single structural breakpoint at `tau = 10`.

```{r scenario2}
n <- 300
tau <- runif(n, 0, 20)

# Create an extreme propensity score distribution
pi_trt <- plogis(-2 + 0.2 * tau)
Trt <- rbinom(n, 1, pi_trt)

# Outcome with a negative shift (slope change) at tau = 10
Y_prob <- plogis(0.5 + 0.2 * tau + 2 * Trt - 0.5 * pmax(0, tau - 10))
Y <- rbinom(n, 1, Y_prob)
dat2 <- data.frame(id = 1:n, tau = tau, Trt = Trt, Y = Y)

fit2 <- bjlm_model() |>
  propensity(Trt ~ tau, data = dat2) |>
  # deltas, omega, rho specified to fit a breakpoint model
  outcome(
    Y ~ tau, 
    b0 = ~ 1 + Trt, 
    b1 = ~ 1, 
    deltas = list(~ 1), 
    omega = list(~ 1), 
    rho = list(~ 1), 
    data = dat2, 
    family = "binomial"
  ) |>
  compile() |>
  fit(iter = 500, warmup = 200, chains = 2, cores = 2, max_weight = 10)

# Since this is a binomial model, tab_bjlm should exponentiate coefficients (Odds Ratios)
tab_bjlm(fit2)
```

---

## Scenario 3: Negative Binomial with Random Intercepts & Latent GP

This is the most complex specification. We simulate longitudinal data where a confounder `X_obs` is observed with noise but enters the outcome and propensity models. We use a Latent Gaussian Process to model `X_obs` smoothly over time. The outcome follows a Negative Binomial distribution with subject-level random intercepts.

```{r scenario3}
n_subj <- 50
n_obs <- 5
tau_out <- runif(n_subj * n_obs, 0, 15)
id_out <- rep(1:n_subj, each = n_obs)

# True smooth confounder
X_true <- sin(tau_out / 3)
Trt_expanded <- rbinom(n_subj * n_obs, 1, plogis(X_true))

# NB outcome with random intercept per subject
mu <- exp(0.5 + 0.1 * tau_out + Trt_expanded + X_true + rnorm(n_subj)[id_out])
Y_nb <- rnbinom(n_subj * n_obs, mu = mu, size = 2)

dat3 <- data.frame(
  id = factor(id_out), 
  tau = tau_out, 
  Trt = Trt_expanded, 
  Y = Y_nb, 
  # Noisy observation of the true confounder
  X_obs = X_true + rnorm(n_subj * n_obs, 0, 0.2)
)

# Sub-sample the first observation per subject for the cross-sectional propensity model
dat3_prop <- dat3[!duplicated(dat3$id), ]

fit3 <- bjlm_model() |>
  propensity(Trt ~ X_obs, data = dat3_prop) |>
  outcome(
    Y ~ tau, 
    b0 = ~ 1 + Trt + X_obs + (1 | id), 
    b1 = ~ 1, 
    data = dat3, 
    family = "negative_binomial"
  ) |>
  latent_gp(
    name = "X_obs", 
    data = dat3, 
    time_var = "tau", 
    obs_var = "X_obs", 
    subject = "id", 
    time_out_var = "tau", 
    time_trt_var = "tau"
  ) |>
  compile() |>
  # The GP sampler relies on the Gamma approximation fix for numerical stability
  fit(iter = 300, warmup = 150, chains = 2, cores = 2)

# Since this is a negative binomial model, tab_bjlm should output Incident Rate Ratios (IRR)
# The summary should also include the `r` parameter (dispersion).
tab_bjlm(fit3)
```
