---
title: "Calibrating a Climate Attribution Model with SBC"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Calibrating a Climate Attribution Model with SBC}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
  # Not evaluated at build time: this walkthrough fits a full latent-GP +
  # change-point model and then runs simulation-based calibration, which refits
  # the model ~100 times per panel -- far beyond the CRAN vignette budget. The
  # code is shown for reference; set eval = TRUE to run it locally.
  eval = FALSE)
```

## The attribution problem

Suppose you have temperature records from a set of stations and you want to know
whether an external **forcing event**, a volcanic eruption, an emissions change,
the onset of a regional regime shift, changed the *rate* of warming, and if so,
when. The clean way to state this is a continuous piecewise (change-point) trend:
a baseline slope, then a slope **change** `delta` at an event time `omega`.

The complication is **natural variability**. Temperature also wanders under slow
internal modes (an oceanic oscillation, say) that have nothing to do with the
forcing. If you ignore it, a run of naturally warm years right after the event
will masquerade as a forced acceleration (a false positive). `bjlm` models that
natural variability explicitly as a latent **Gaussian process** (GP), informed
by a noisily observed index (here an oscillation index `enso`), and estimates the
forced change-point *jointly* with it.

That joint estimation raises the question this vignette is about:

> **Does the GP soak up systematic (forced) variation from the outcome model, or
> vice versa?** If the GP is too flexible it can absorb the forced trend shift,
> hiding a real acceleration. If it is too rigid, genuine natural variability
> leaks into the change-point, manufacturing an acceleration that is not there.

Neither failure is visible from a single fit, which always returns *some* split
between "forced" and "natural." The tool that detects it is **simulation-based
calibration** (SBC): if the sampler and priors recover the truth on data
simulated from the model, then the `delta` (attribution) posterior and the GP
posterior are each honest, and in particular are not trading variance between
them. We build the model, then certify exactly that.

## 1. A generative model for forced trend + natural variability

We work in **decades since 1970** so the time axis runs over `[0, 5]` (1970 to
2020); this keeps the change-point priors on a natural scale. Each station sees
the *same* forced response but its *own* natural-variability trajectory.

```{r simulate}
library(bjlm)
set.seed(1)

n_series <- 16      # stations / ensemble members
n_obs    <- 12      # annual-ish observations per station
series   <- rep(seq_len(n_series), each = n_obs)
decade   <- rep(seq(0, 5, length.out = n_obs), n_series)   # decades since 1970
sig      <- function(x) 1 / (1 + exp(-x))

## --- true parameters ---
alpha_true <- 0.4;  rho_gp_true <- 1.2;  sigx_true <- 0.12   # natural variability (GP)
b0_true    <- 0.0;  bgp_true    <- 1.0                       # baseline + GP loading
b1_true    <- 0.05                                           # baseline warming (per decade)
delta_true <- 0.20                                           # FORCED slope change
omega_true <- 3.0                                            # event at ~year 2000
rhocp_true <- 4.0                                            # transition sharpness
sigma_true <- 0.08                                           # measurement noise

## --- draw one natural-variability path per station and build temperatures ---
grid <- sort(unique(decade))
D    <- as.matrix(dist(grid))
K    <- alpha_true^2 * exp(-0.5 * (D / rho_gp_true)^2) + diag(1e-6, n_obs)
L    <- t(chol(K))

temp <- numeric(length(series)); enso <- numeric(length(series))
for (s in seq_len(n_series)) {
  idx <- which(series == s); ts <- decade[idx]
  f   <- as.numeric(L %*% rnorm(n_obs))          # latent natural variability
  enso[idx] <- f + rnorm(n_obs, 0, sigx_true)    # noisily observed oscillation index
  d   <- ts - omega_true
  mu  <- b0_true + bgp_true * f +                # level + natural variability
         b1_true * (ts - omega_true) +           # baseline warming
         delta_true * d * sig(rhocp_true * d)    # FORCED slope change after the event
  temp[idx] <- mu + rnorm(n_obs, 0, sigma_true)
}

dat <- data.frame(temp = temp, decade = decade, enso = enso, series = factor(series))
```

Each generative term maps to a model parameter: `b1` is the baseline warming
slope, `delta` is the forced change in slope (the attribution target), `omega`
the event time, `rho` its abruptness, and the GP contributes `bgp * f` to the
temperature level, where `f` is the latent oscillation with amplitude `alpha`,
lengthscale `rho_gp`, observed through `enso` with noise `sigx`.

## 2. Fitting the joint model

The latent GP is declared on the observed `enso` index and loads on the outcome
intercept (`b0 = ~ 1 + enso`); the change-point lives in the `deltas`/`omega`/`rho`
blocks. No propensity model is needed, this is a calibration/attribution problem,
not a treatment-effect problem, so `bjlm` fits a true posterior.

```{r fit}
clim_model <- bjlm_model() |>
  outcome(
    temp ~ decade,
    b0     = ~ 1 + enso,     # GP (natural variability) loads on the level
    b1     = ~ 1,            # baseline warming slope
    deltas = list(~ 1),      # FORCED slope change
    omega  = list(~ 1),      # event time
    rho    = list(~ 1),      # transition sharpness
    data   = dat
  ) |>
  latent_gp(
    name = "enso", data = dat, obs_var = "enso",
    time_var = "decade", time_out_var = "decade", time_trt_var = "decade",
    subject = "series", kernel = "se"
  ) |>
  compile()

# Priors on the year/decade scale: the event lies within the record, the
# transition is resolvable, and the forced slope change is weakly informative.
clim_priors <- bjlm_priors(outcome = smoothbp_priors(
  b0     = list("(Intercept)" = prior_normal(0, 1), "enso" = prior_normal(0, 1)),
  b1     = prior_normal(0, 0.3),
  deltas = prior_normal(0, 0.3),
  omega  = prior_normal(3, 1, lb = 0, ub = 5),
  rho    = prior_normal(4, 1.5, lb = 0),
  sigma  = prior_invgamma(3, 2)))

clim_fit <- fit(clim_model, priors = clim_priors, chains = 4, iter = 2000, cores = 2)

summary(clim_fit, model = "outcome")
```

`delta1_(Intercept)` is the attributed change in warming rate. `plot_gp()` shows
the natural-variability paths the model peeled off, and `plot_predictions()` the
reconstructed temperature trajectories:

```{r plots}
plot_gp(clim_fit, n_subjects = 4)
plot_predictions(clim_fit, type = "subject", n_subjects = 4)
```

A single fit like this always produces an answer. The next section asks whether
that answer can be *trusted*, and specifically whether the forced and natural
parts are cleanly separated.

## 3. Certifying the decomposition with SBC

`sbc()` re-runs the whole "draw truth from the prior, simulate, re-fit, rank the
truth in its posterior" loop many times. For a calibrated model the ranks are
uniform. The quantities that matter here are checked by default:

- **`delta1_(Intercept)`** — the forced slope change (the attribution claim);
- **`b1_(Intercept)`**, **`omega1_(Intercept)`**, **`rho1_(Intercept)`** — the rest of the trend;
- **`enso_alpha`**, **`enso_rho`**, **`b0_enso`** — the GP amplitude, lengthscale, and loading.

If `delta` *and* the GP quantities are all calibrated **at the same time**, then
neither component is systematically borrowing the other's variance, which is
exactly the confounding we were worried about.

```{r sbc}
cal <- sbc(clim_model, priors = clim_priors, reps = 100, iter = 1500, chains = 2, cores = 2)

cal          # plain-language verdict per quantity
plot(cal)    # rank histograms; flat within the grey band = calibrated
```

Read the printed table directly. A verdict of *"well calibrated"* on
`delta1_(Intercept)` means the attributed acceleration is neither inflated nor
deflated by the presence of the GP. Watch the `n_used` column too: if `omega` is
identified in only a fraction of prior-drawn datasets, that is telling you the
event time is often unrecoverable when the forcing is weak, useful to know before
you make a claim about *when* the change happened.

Because the natural-variability confounding is the point, it is worth adding the
two directional functionals explicitly, the pre- and post-event slopes, so you
can see that both the baseline and the accelerated regime are calibrated:

```{r sbc-functionals}
cal2 <- sbc(clim_model, priors = clim_priors, reps = 100,
  functionals = list(
    slope_pre  = function(p) p[["b1_(Intercept)"]],
    slope_post = function(p) p[["b1_(Intercept)"]] + p[["delta1_(Intercept)"]]
  ))
cal2
```

## 4. How much prior does the separation need? A lengthscale sweep

The single knob that governs how much systematic variation the GP *can* absorb is
its **lengthscale prior**. A very diffuse lengthscale prior lets the GP wander on
long time scales, precisely the scales on which a forced trend lives, so it can
quietly eat the acceleration and leave `delta` mis-calibrated. `bjlm` lets you
find that breaking point directly:

```{r prior-sensitivity}
sens <- sbc_prior_sensitivity(
  clim_model,
  grid = c(0.3, 0.6, 1.0, 1.5, 2.0),   # GP lengthscale prior widths (log-sd)
  target = "^delta",                   # watch the attribution functional
  reps = 100, iter = 1500, chains = 2, cores = 2
)

sens         # min slope-calibration p-value at each prior width
plot(sens)   # p-value (y) vs prior width (x); the red line is p = 0.05
```

Read the plot along the width axis, left to right. At **narrow** lengthscale
priors the slope-calibration p-value sits **above** the red p = 0.05 line: the
GP is constrained, and the trend-versus-GP decomposition is trustworthy. As the
prior **widens**, the p-value falls and eventually drops **below** the line
(here at a width of about 1): the GP now has enough freedom to absorb the forced
trend, so a "forced change" is partly just re-labelled natural variability. The
`print` output names that crossing width for you.

That crossing point, the widest lengthscale prior your data can tolerate before
the attribution stops being trustworthy, is a concrete answer to how much prior
information your decomposition requires. It is a statement about *your* model on
*your* observation grid, and not something a generic calibration check can give
you.

## Takeaways

- A latent GP and a continuous change-point can, in principle, trade variance:
  the GP can hide a forced trend, or absorb none and let natural variability
  fake one. Which way it goes depends on the priors and the sampling grid.
- `sbc()` certifies both directions at once. When `delta` (the attribution) and
  the GP amplitude/lengthscale/loading are *jointly* calibrated, the
  decomposition is honest.
- `sbc_prior_sensitivity()` locates the lengthscale prior width at which that
  honesty breaks, giving you a defensible statement of how much prior information
  your attribution actually requires.
