---
title: "Getting Started with bjlm"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with bjlm}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

## Overview

**`bjlm`** (Bayesian Joint Longitudinal Modelling) is a standalone, high-performance R package designed for joint longitudinal and causal analysis. It simultaneously estimates:
1. A **Propensity Score Model** ($p(Trt \mid X)$) capturing subject-level treatment allocation.
2. A **Weighted Outcome Model** ($p(Y \mid \tau, \text{Trt}, \dots)$) incorporating subject-specific inverse probability weights (IPW) to correct for confounding.

By performing joint estimation, `bjlm` automatically propagates the uncertainty of the propensity weights into the causal effect posterior, preventing the underestimation of credible intervals common in two-stage statistical methods. Under the hood, a fast, multithreaded sampler written in Rust handles the MCMC execution.

---

## The Next-Generation Piped API

The package features an elegant, block-based declarative builder API that leverages R's native pipe operator (`|>`). It isolates treatment, GP, and outcome models dynamically into individual blocks:

```R
library(bjlm)

fit <- bjlm_model() |>
  propensity(Trt ~ X1 + X2, data = df_subject) |>
  outcome(Y ~ X1 + Trt, data = df_obs) |>
  compile() |>
  fit()
```

### Pipelines and Modular Steps

* **`bjlm_model()`**: Initializes an empty model specification builder.
* **`propensity(model, formula, data, family = binomial("logit"))`**: Stages the treatment assignment model (LHS must be binary `0`/`1`).
* **`outcome(model, formula, b0, b1, deltas, omega, rho, data, family = gaussian("identity"))`**: Stages the outcome model. Can fit standard linear regressions or piecewise change-point models.
* **`latent_gp(model, name, time, subject, kernel = "se")`**: Optionally registers continuous-time latent Gaussian Process confounders (can add multiple GPs sequentially).
* **`compile(model)`**: Performs alignment across subject-level and observation-level data, validates GP structures, and enforces the **strict Bayesian Cut** (which prevents outcome feedback into the propensity model).
* **`fit(compiled_model, priors = NULL, ...)`**: Compiles priors and runs the fast Rust sampler.

---

## 1. Quick Start: Standard Regression (Zero-Breakpoint Speed Shortcut)

If you wish to fit a standard linear model without any breakpoints, simply omit the piecewise parameters (`b0`, `b1`, etc.) from the `outcome()` block. The compiler automatically triggers the **zero-breakpoint speed shortcut**, which bypasses sigmoid calculations entirely for maximum performance.

### Simulating Confounded Data

Here, we simulate cross-sectional data where treatment `Trt` is confounded by `X1`:

```{r simulate-data}
set.seed(42)
n <- 300

# Confounder X1
X1 <- rnorm(n)

# Confounded treatment probability
pi_true <- plogis(0.5 + 0.8 * X1)
Trt <- rbinom(n, 1, pi_true)

# Outcome confounded by X1, with true treatment effect (ATE) of 2.0
Y <- 5.0 + 2.0 * Trt + 1.5 * X1 + rnorm(n)

dat <- data.frame(Y = Y, Trt = Trt, X1 = X1)
```

### Specifying and Fitting the Model

```{r fit-linear}
# Build and compile the joint model
linear_model <- bjlm_model() |>
  propensity(Trt ~ X1, data = dat) |>
  outcome(Y ~ X1 + Trt, data = dat) |>
  compile()

# Fit using the fast Rust sampler
fit_linear <- fit(linear_model, chains = 2L, iter = 1000L, warmup = 500L)
```

---

## 2. Advanced: Piecewise Change-Point Models

For repeated-measures or longitudinal data where the trajectory changes slope at one or more unknown time thresholds, `bjlm` supports smoothed piecewise change-point models:

$$
\mu_i = b0_i + b1_i(\tau_i - \omega_{1i}) + \sum_{k=1}^{K} \delta_{ki}(\tau_i - \omega_{ki})\, \sigma\!\left[(\tau_i - \omega_{ki})\rho_{ki}\right]
$$

where:
* **$\tau$** is the time variable.
* **$b0$** is the level at the first breakpoint.
* **$b1$** is the pre-change slope.
* **$\omega_k$** represents the location of the $k$-th change-point.
* **$\delta_k$** is the change in slope at change-point $k$.
* **$\rho_k$** is the transition sharpness.

### Separate Datasets for Propensity and Outcome Models

In longitudinal joint modeling, treatment allocation and propensity score covariates are typically **subject-level** properties (constant over time for each subject), while outcomes and time values are **observation-level** properties (varying over time). 

A key strength of `bjlm` is that you do not need to repeat subject-level characteristics across observation rows. You can pass a subject-level data frame (one row per subject) to `propensity()`, and an observation-level data frame (multiple rows per subject) to `outcome()`. When you run `compile()`, `bjlm` automatically groups and aligns these datasets using the grouping variable specified in the outcome random intercepts (e.g. `(1 | id)`).

Let's simulate this structure:

```{r simulate-piecewise}
# 1. Subject-level dataset (constant covariates and treatment)
set.seed(42)
n_subj <- 30
id <- factor(1:n_subj)
X1 <- rnorm(n_subj)
pi_subj <- plogis(0.2 + 0.8 * X1)
Trt <- rbinom(n_subj, 1, pi_subj)

dat_subjects <- data.frame(id = id, X1 = X1, Trt = Trt)

# 2. Observation-level dataset (repeated measures, 5 timepoints per subject)
n_obs_per_subj <- 5
tau_values <- seq(0, 4, length.out = n_obs_per_subj)

obs_list <- lapply(1:n_subj, function(i) {
  subj_id  <- id[i]
  subj_trt <- Trt[i]
  subj_x1  <- X1[i]
  
  # Piecewise outcome trajectory with a breakpoint at omega = 2.0
  # Level intercept includes subject-specific random intercept u_i (SD = 0.5)
  u_i     <- rnorm(1, mean = 0, sd = 0.5)
  b0_i    <- 4.0 + 1.5 * subj_trt + 1.0 * subj_x1 + u_i
  b1_i    <- -0.2
  delta_i <- 0.8
  omega_i <- 2.0
  rho_i   <- 5.0
  
  Y <- vapply(tau_values, function(t) {
    sig <- 1 / (1 + exp(-(t - omega_i) * rho_i))
    b0_i + b1_i * (t - omega_i) + delta_i * (t - omega_i) * sig + rnorm(1, sd = 0.2)
  }, double(1))
  
  data.frame(
    id  = subj_id,
    tau = tau_values,
    Y   = Y
  )
})

dat_observations <- do.call(rbind, obs_list)
```

### Specifying and Fitting the Longitudinal Model

Now we build and compile the model, specifying the subject-level data frame for the propensity stage and the observation-level repeated measures data frame for the outcome stage:

```{r fit-piecewise}
# Build and compile the joint model
pw_model <- bjlm_model() |>
  propensity(Trt ~ X1, data = dat_subjects) |>
  outcome(
    formula = Y ~ tau,               # Outcome Y and time variable tau
    b0      = ~ 1 + Trt + (1 | id),  # Level with subject random intercept
    b1      = ~ 1,                   # Initial slope
    deltas  = list(~ 1),             # Slope change parameter
    omega   = list(~ 1),             # Unknown change-point location
    rho     = list(~ 1),             # Transition sharpness
    data    = dat_observations
  ) |>
  compile()

# Fit using the fast Rust sampler
fit_pw <- fit(pw_model, chains = 4L, iter = 2000L, warmup = 1000L, cores = 4L)
```

---

## 3. Post-Processing & Diagnostics

`bjlm` includes a premium suite of post-processing and diagnostic methods tailored to joint longitudinal estimation:

### A. Print and Summary Methods
The `summary()` method extracts MCMC stats as a structured data frame, while `print()` produces a formatted terminal layout.

```{r postprocess}
# Print high-level diagnostics and metadata
print(fit_linear)

# Obtain posterior means, SDs, and Rhat values
summary(fit_linear, model = "both")
```

### B. Weight Diagnostics
Since weights are calculated at every MCMC iteration, you can inspect stabilised weight distributions and max trimming thresholds directly:

```{r weight-diagnostics}
weight_diagnostics(fit_linear)
```

### C. Causal Effects
Extract and print average treatment effects (ATE) or treatment effects on the treated (ATT):

```{r causal-effect}
# Get credible intervals for the treatment coefficient
causal_effect(fit_linear, param = "b0_Trt")
```

### D. Trace and Density Plots
Inspect MCMC chain convergence and parameter densities:

```{r plotting}
# Plot posterior trace-lines and density distributions
plot(fit_linear)
```

---

## 4. Side-by-Side Model Comparison Tables

The package provides `tab_bjlm()` to generate professional, CRAN-standard fixed-effects comparison tables across multiple fitted models. It aligns parameters across rows and displays models side-by-side:

```{r compare-table}
# Generate comparison table
comparison_table <- tab_bjlm(
  fit_linear, 
  fit_pw, 
  labels = c("Linear IPW Model", "Piecewise IPW Model")
)

# If the 'gt' package is installed, this returns a beautiful gt table.
# Otherwise, it falls back to a clean knitr Markdown table.
print(comparison_table)
```
