---
title: "Getting Started with bjlm"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with bjlm}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
  # Not evaluated at build time: this walkthrough fits several full MCMC models
  # (including a latent-GP model) and runs lfo_cv(), which refits repeatedly --
  # well beyond the CRAN vignette time budget. The code is shown for reference;
  # set eval = TRUE to run it locally.
  eval = FALSE)
```

## Overview

**`bjlm`** (Bayesian Joint Longitudinal Modelling) is a high-performance R package designed for joint longitudinal and causal analysis. It natively resolves one of the most challenging problems in observational data: **time-mismatched observations**. 

Often, covariates, treatment allocations, and outcomes are measured at entirely different time points. `bjlm` handles this seamlessly by modeling time-varying confounders as continuous-time **Gaussian Processes (GP)**. It simultaneously estimates:
1. A **Propensity Score Model** ($p(Trt \mid X_{gp})$) evaluating the probability of treatment based on the interpolated GP state of the covariates at the exact time of treatment.
2. A **Weighted Outcome Model** ($p(Y \mid \tau, \text{Trt}, \dots)$) incorporating subject-specific inverse probability weights (IPW) to correct for confounding.

By performing joint estimation, `bjlm` automatically propagates the uncertainty of the missing data and the propensity weights into the causal effect posterior, preventing the underestimation of credible intervals common in two-stage statistical methods. Under the hood, a fast, multithreaded sampler written in Rust handles the MCMC execution.

---

## The Next-Generation Piped API

The package features an elegant, block-based declarative builder API that leverages R's native pipe operator (`|>`):

```R
fit <- bjlm_model() |>
  propensity(Trt ~ X_gp, data = dat_treatment) |>
  latent_gp(name = "X_gp", time = "tau", subject = "id", kernel = "se") |>
  outcome(Y ~ tau + Trt, data = dat_outcome) |>
  compile() |>
  fit()
```

### Pipelines and Modular Steps

* **`bjlm_model()`**: Initializes an empty model specification builder.
* **`propensity(model, formula, data, family)`**: Stages the treatment assignment model. Use the default `family = binomial("logit")` for binary treatments, or `family = gaussian("identity")` for continuous treatments (generalized propensity score weighting).
* **`latent_gp(model, name, data, obs_var, time_var, time_trt_var, time_out_var, subject, kernel = "se")`**: Registers continuous-time latent Gaussian Process confounders.
* **`outcome(model, formula, b0, b1, deltas, omega, rho, data, family)`**: Stages the outcome model. Can fit standard linear regressions or piecewise change-point models, with `family` supporting `"gaussian"` (default), `"binomial"`, or `"negative_binomial"` outcomes (the latter mixes cleanly with `latent_gp()` confounders via a conditional-transport sampler).
* **`compile(model)`**: Validates GP structures, aligns time-points, and enforces the strict Bayesian Cut to prevent outcome-to-propensity feedback.
* **`fit(compiled_model, ...)`**: Compiles priors and runs the fast Rust sampler.

---

## 1. Quick Start: Standard Cross-Sectional IPW

For simple cross-sectional data where variables are measured concurrently, `bjlm` acts as a fast Marginal Structural Model (MSM) / Inverse Probability Weighting (IPW) estimator. You can omit the `latent_gp` block entirely:

```r
# Simple standard linear regression shortcut (cross-sectional)
basic_model <- bjlm_model() |>
  propensity(Trt ~ X1 + X2, data = df_cross_sectional) |>
  outcome(Y ~ X1 + Trt, data = df_cross_sectional) |>
  compile()

fit_basic <- fit(basic_model, chains = 2, iter = 1000)
```

---

## 2. Joint Longitudinal Modelling with Continuous-Time Confounders

In realistic clinical or observational data, time-varying covariates, treatment allocation, and outcomes rarely align. A patient might have a lab test on Tuesday, receive treatment on Thursday, and have an outcome measured next Monday.

### Why use a Gaussian Process instead of Linear Interpolation?
While it may seem simpler to just linearly interpolate a covariate between its observed time points, linear interpolation assumes we know the unobserved values with 100% certainty. In reality, medical and physiological covariates can fluctuate non-linearly between measurements. A Gaussian Process solves this by providing a distribution of possible trajectories. Crucially, the GP quantifies **uncertainty** during the periods between observations. Because `bjlm` fits the GP jointly with the propensity and outcome models, this uncertainty naturally propagates through the inverse-probability weights into the final treatment effect estimate. This prevents the model from being over-confident and ensures your credible intervals maintain proper coverage.

`bjlm` uses Gaussian Processes to impute this smooth, subject-specific trajectory for the confounder. The propensity model then queries this trajectory at the exact time of treatment.

### Simulating Time-Mismatched Longitudinal Data

Let's simulate a scenario where a time-varying covariate $X$, treatment $Trt$, and outcome $Y$ are measured at different, mismatched time points:

```{r simulate-mismatched}
library(bjlm)
set.seed(42)
n_subj <- 30

# 1. Simulate the GP confounder (Covariate X measured at random times)
tau_cov <- runif(n_subj * 4, 0, 10)
id_cov  <- rep(1:n_subj, each = 4)
# Subject-specific smooth trajectories
X_obs <- sin(tau_cov / 2) + rnorm(n_subj * 4, sd = 0.2)
dat_covariates <- data.frame(id = factor(id_cov), tau = tau_cov, X_obs = X_obs)

# 2. Simulate Treatment Allocation (Measured at a single specific time per subject)
tau_trt <- runif(n_subj, 2, 8)
id_trt  <- 1:n_subj
# True unobserved confounder state at treatment time
X_true_trt <- sin(tau_trt / 2)
# Treatment probability depends on the time-varying confounder
pi_trt <- plogis(-0.5 + 2.0 * X_true_trt)
Trt <- rbinom(n_subj, 1, pi_trt)
dat_treatment <- data.frame(id = factor(id_trt), tau_trt = tau_trt, Trt = Trt, X_gp = 0) # X_gp placeholder

# 3. Simulate Outcome (Measured longitudinally at yet another set of times)
tau_out <- runif(n_subj * 5, 0, 10)
id_out  <- rep(1:n_subj, each = 5)
# True unobserved confounder state at outcome time
X_true_out <- sin(tau_out / 2)
# Outcome confounded by Trt and X
Y <- 2.0 + 1.5 * rep(Trt, each = 5) + 0.8 * X_true_out + rnorm(n_subj * 5, sd = 0.5)
dat_outcome <- data.frame(id = factor(id_out), tau_out = tau_out, Y = Y, X_gp = 0) # X_gp placeholder
```

### Specifying and Fitting the Model

We construct a joint model where `X_obs` informs the `latent_gp`. The GP generates a continuous state `X_gp` which is dynamically queried by the propensity and outcome formulas. 

```{r fit-gp-model}
gp_model <- bjlm_model() |>
  # Treatment block queries X_gp at tau_trt
  propensity(Trt ~ X_gp, data = dat_treatment) |>
  
  # Declare the GP for X_obs across time tau
  latent_gp(
    name = "X_gp",
    data = dat_covariates,
    time_var = "tau",
    obs_var = "X_obs",
    time_trt_var = "tau_trt",
    time_out_var = "tau_out",
    subject = "id",
    kernel = "se"
  ) |>
  
  # Outcome block queries X_gp at tau_out. 
  # We include a random subject intercept (1 | id).
  outcome(
    formula = Y ~ tau_out,
    b0 = ~ 1 + Trt + X_gp + (1 | id),
    b1 = ~ 1,
    data = dat_outcome
  ) |>
  compile()

# Fit using the fast Rust sampler
fit_long <- fit(gp_model, chains = 4, iter = 2000, cores = 2)
```

---

## 3. Post-Processing & Diagnostics

`bjlm` includes a premium suite of post-processing and diagnostic methods tailored to joint longitudinal estimation, allowing you to visualise everything from raw data fits to unobserved confounder trajectories.

### Print and Summary

The `print()` method gives a terminal-friendly overview, while `summary()` provides posterior mean, standard deviation, and R-hat statistics.

```{r summary}
print(fit_long)
summary(fit_long, model = "both")
```

### Predictive Fit

Extract point estimates and credible intervals for the fitted outcome trajectory using `fitted()`.

```{r fitted}
fitted_draws <- fitted(fit_long, summary = TRUE)
head(fitted_draws)
```

### Visualising Model Predictions

Use `plot_predictions()` to overlay posterior fitted trajectories (with 95% credible intervals) on top of the raw data. 

```{r plot-predictions}
# Plot population-level or subject-specific fits
plot_predictions(fit_long, type = "subject", n_subjects = 4)
```

### Visualising Gaussian Processes

The true power of `bjlm` is the continuous-time imputation. You can visualise the inferred, unobserved GP trajectories for individual subjects using `plot_gp()`:

```{r plot-gp}
# Inspect the inferred continuous-time confounder paths
plot_gp(fit_long, n_subjects = 4)
```

### Propensity and Weight Diagnostics

Understand treatment assignment overlap and inverse probability weight distributions using `plot_propensity()`.

```{r plot-propensity}
# Density overlap of propensity scores and distribution of weights
plot_propensity(fit_long, type = "both")
```

---

## 4. Leave-Future-Out Cross Validation

For longitudinal models, standard leave-one-out cross validation often underestimates true predictive error because it attempts to predict past data using future data.

`bjlm` natively supports **Leave-Future-Out Cross Validation (LFO-CV)**. By sequentially predicting time $t+1$ using only data up to time $t$, LFO-CV provides a rigorous assessment of the model's forward-predictive capability. It uses Pareto Smoothed Importance Sampling (PSIS) to approximate the refits, making it incredibly fast.

```{r lfo-cv}
# Perform LFO-CV, starting the forward-validation from time tau = 2.0
lfo_res <- lfo_cv(fit_long, min_tau = 2.0)

# Print the Expected Log Predictive Density (ELPD) summary
print(lfo_res)

# Plot the PSIS Pareto k diagnostic values to assess reliability
plot(lfo_res)
```

High Pareto $k$ values ($k > 0.7$) in the diagnostic plot indicate points where the importance sampling approximation struggles, but `lfo_cv()` will automatically issue warnings when this occurs so you can trust the results.

### Comparing Models with LFO-CV

LFO-CV is especially useful for model selection. Let's fit a competing model that incorrectly omits the treatment effect from the outcome block, and see if LFO-CV correctly identifies the true model (which includes the treatment effect) as having superior predictive performance.

```{r compare-models}
# Fit a competing model omitting the Treatment effect from the outcome
bad_model <- bjlm_model() |>
  propensity(Trt ~ X_gp, data = dat_treatment) |>
  latent_gp(
    name = "X_gp",
    data = dat_covariates,
    time_var = "tau",
    obs_var = "X_obs",
    time_trt_var = "tau_trt",
    time_out_var = "tau_out",
    subject = "id",
    kernel = "se"
  ) |>
  outcome(
    formula = Y ~ tau_out,
    b0 = ~ 1 + X_gp + (1 | id), # Incorrectly omitting Trt
    b1 = ~ 1,
    data = dat_outcome
  ) |>
  compile()

fit_bad <- fit(bad_model, chains = 4, iter = 2000, cores = 2)

# Run LFO-CV on the competing model
lfo_bad <- lfo_cv(fit_bad, min_tau = 2.0)

# Formally compare the true model and the bad model using the loo package:
loo::loo_compare(list(True_Model = lfo_res, Bad_Model = lfo_bad))
```

As expected, the true model (which correctly specifies the outcome generating process) will yield a significantly higher ELPD score, demonstrating LFO-CV's ability to select the better forward-predicting model!

---

## 5. Checking Calibration with Simulation-Based Calibration (SBC)

Before you trust a model's estimates, it is worth asking a sharper question than "did it converge?": **does this sampler, with these priors, actually recover the truth on data that look like mine?** Simulation-based calibration (SBC; Talts et al. 2018) answers it directly. For many replicates it draws a full parameter set from the prior, simulates a dataset from it, re-fits, and records where each true value falls within its posterior. If the model is calibrated, those ranks are uniform; a systematic departure tells you the intervals are too narrow (overconfident), too wide, or biased.

`bjlm` builds this in with `sbc()`. It is designed so the result is easy to read, so you are never left staring at rank histograms wondering what they mean.

### Running SBC

You can call `sbc()` on a **compiled model** (to check calibration *before* committing to an expensive fit) or on a **fitted model** (which reuses that fit's model, priors, and design). Calibration is cleanest for a change-point outcome model, where the interpretable quantities, the segment slopes and breakpoints, are exactly what SBC checks by default.

Note that SBC only needs the model's *design* (its covariates, times, and subject structure), not its response: each replicate simulates a fresh `Y`. So a small template dataset with the right columns is enough to compile the model we want to check:

```{r sbc-data}
set.seed(1)
n_subj <- 25; n_obs <- 6
subj  <- rep(seq_len(n_subj), each = n_obs)
dat_cp <- data.frame(
  Y   = 0,                                              # placeholder; SBC simulates Y
  tau = rep(seq(0, 6, length.out = n_obs), n_subj),     # change-point stays inside [0, 6]
  Trt = rbinom(n_subj, 1, 0.5)[subj],
  id  = factor(subj)
)
```

```{r sbc-basic}
# A change-point outcome model: does the treatment change the slope, and when?
cp_model <- bjlm_model() |>
  outcome(
    Y ~ tau,
    b0     = ~ 1,
    b1     = ~ 1,              # initial slope
    deltas = list(~ 1 + Trt),  # slope CHANGE (modified by treatment) = "did it accelerate?"
    omega  = list(~ 1),        # breakpoint location
    rho    = list(~ 1),        # transition sharpness
    data   = dat_cp
  ) |>
  compile()

# Use priors that keep the change-point inside the observed time range; a
# breakpoint prior with mass outside [min(tau), max(tau)] is unidentified for
# reasons unrelated to the sampler, and sbc() will warn you if so.
cal_priors <- bjlm_priors(outcome = smoothbp_priors(
  b0 = prior_normal(0, 3), b1 = prior_normal(0, 1), deltas = prior_normal(0, 1),
  omega = prior_normal(3, 1, lb = 0), rho = prior_normal(4, 1, lb = 0),
  sigma = prior_invgamma(3, 2)))

cal <- sbc(cp_model, priors = cal_priors, reps = 100, iter = 1500, chains = 2)

cal          # a plain-language verdict per quantity
plot(cal)    # rank histograms with the uniform envelope band
```

The `print()` output is the point of the whole exercise: each quantity gets a one-line reading, for example *"well calibrated"*, *"intervals too NARROW (overconfident)"*, or *"estimates biased HIGH"*, so you do not need to decode histogram shapes yourself. `plot()` shows the underlying rank histograms; a calibrated quantity is flat within the shaded band.

To run SBC on a model you have already fitted, just pass the fit; it reuses the same model, priors, and design automatically:

```{r sbc-on-fit}
cal_fit <- sbc(fit_long, reps = 100)
```

### Reading the guards

Two robustness details are handled for you and reported honestly:

* **Thinning to independence.** SBC assumes independent draws, so `sbc()` thins each quantity by its effective sample size before ranking.
* **Convergence filtering.** A replicate whose fit did not converge would otherwise masquerade as miscalibration, so `sbc()` drops per-quantity draws with high R-hat and reports an `n_used` column. A low `n_used` for, say, `omega1_(Intercept)` is itself informative: it means the change-point is often *non-identified* on prior-drawn data. The overall failed/degenerate-fit rate is reported too.

Because SBC is conditional on convergence, and because IPW-weighted (marginal structural) fits target a pseudo-posterior, `sbc()` prints a short note when either applies so you can interpret the result correctly.

### Custom target quantities

The defaults, the initial slope, each slope difference, the breakpoints and sharpness, the GP amplitude and lengthscale, and the scales, cover most needs. You can add your own functionals as closures of the named parameter vector. For example, to calibrate the *treated* group's total slope change ("did the treated group accelerate?"):

```{r sbc-functionals}
cal2 <- sbc(cp_model, priors = cal_priors, reps = 100,
  functionals = list(
    treated_accel = function(p) p[["delta1_(Intercept)"]] + p[["delta1_Trt"]]
  ))
```

### Spike-and-slab models

When you fit with `prior_spike_slab()`, the selected coefficients have a point mass at zero, so a rank statistic is ill-defined for them. `sbc()` automatically excludes them from the rank check (calibrating the always-continuous quantities instead); calibrate the selection itself with `pip()`, which reports posterior inclusion probabilities.

### Prior sensitivity: how much prior does my decomposition need?

The distinctive question for a joint trend-plus-GP model is: *how much prior information does my decomposition actually require?* If the GP lengthscale prior is too diffuse, the GP can absorb genuine trend, quietly degrading the calibration of the slope estimates. `sbc_prior_sensitivity()` answers this by sweeping the GP lengthscale prior width and reporting where the slope (`delta`) calibration breaks:

```{r sbc-prior-sensitivity}
sens <- sbc_prior_sensitivity(
  gp_model,
  grid = c(0.3, 0.6, 1.0, 1.5, 2.0),  # GP lengthscale prior widths (sdlog)
  reps = 100
)

sens         # a table of slope-calibration p-value vs prior width
plot(sens)   # below the red line, the slope decomposition is miscalibrated
```

This turns a vague worry ("is my GP eating the trend?") into a concrete threshold, the prior width beyond which you can no longer trust the trend-versus-GP split, which no generic SBC tool can tell you for your model.
