---
title: "Getting Started with bjlm"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with bjlm}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

## Overview

**`bjlm`** (Bayesian Joint Longitudinal Modelling) is a high-performance R package designed for joint longitudinal and causal analysis. It natively resolves one of the most challenging problems in observational data: **time-mismatched observations**. 

Often, covariates, treatment allocations, and outcomes are measured at entirely different time points. `bjlm` handles this seamlessly by modeling time-varying confounders as continuous-time **Gaussian Processes (GP)**. It simultaneously estimates:
1. A **Propensity Score Model** ($p(Trt \mid X_{gp})$) evaluating the probability of treatment based on the interpolated GP state of the covariates at the exact time of treatment.
2. A **Weighted Outcome Model** ($p(Y \mid \tau, \text{Trt}, \dots)$) incorporating subject-specific inverse probability weights (IPW) to correct for confounding.

By performing joint estimation, `bjlm` automatically propagates the uncertainty of the missing data and the propensity weights into the causal effect posterior, preventing the underestimation of credible intervals common in two-stage statistical methods. Under the hood, a fast, multithreaded sampler written in Rust handles the MCMC execution.

---

## The Next-Generation Piped API

The package features an elegant, block-based declarative builder API that leverages R's native pipe operator (`|>`):

```R
library(bjlm)

fit <- bjlm_model() |>
  propensity(Trt ~ X_gp, data = dat_treatment) |>
  latent_gp(name = "X_gp", time = "tau", subject = "id", kernel = "se") |>
  outcome(Y ~ tau + Trt, data = dat_outcome) |>
  compile() |>
  fit()
```

### Pipelines and Modular Steps

* **`bjlm_model()`**: Initializes an empty model specification builder.
* **`propensity(model, formula, data)`**: Stages the treatment assignment model (LHS must be binary `0`/`1`).
* **`latent_gp(model, name, time, subject, kernel = "se")`**: Registers continuous-time latent Gaussian Process confounders.
* **`outcome(model, formula, b0, b1, deltas, omega, rho, data)`**: Stages the outcome model. Can fit standard linear regressions or piecewise change-point models.
* **`compile(model)`**: Validates GP structures, aligns time-points, and enforces the strict Bayesian Cut to prevent outcome-to-propensity feedback.
* **`fit(compiled_model, ...)`**: Compiles priors and runs the fast Rust sampler.

---

## 1. Quick Start: Standard Cross-Sectional IPW

For simple cross-sectional data where variables are measured concurrently, `bjlm` acts as a fast Marginal Structural Model (MSM) / Inverse Probability Weighting (IPW) estimator. You can omit the `latent_gp` block entirely:

```r
# Simple standard linear regression shortcut (cross-sectional)
basic_model <- bjlm_model() |>
  propensity(Trt ~ X1 + X2, data = df_cross_sectional) |>
  outcome(Y ~ X1 + Trt, data = df_cross_sectional) |>
  compile()

fit_basic <- fit(basic_model, chains = 2, iter = 1000)
```

---

## 2. Joint Longitudinal Modelling with Continuous-Time Confounders

In realistic clinical or observational data, time-varying covariates, treatment allocation, and outcomes rarely align. A patient might have a lab test on Tuesday, receive treatment on Thursday, and have an outcome measured next Monday.

`bjlm` uses Gaussian Processes to impute a smooth, subject-specific trajectory for the confounder. The propensity model then queries this trajectory at the exact time of treatment.

### Simulating Time-Mismatched Longitudinal Data

Let's simulate a scenario where a time-varying covariate $X$, treatment $Trt$, and outcome $Y$ are measured at different, mismatched time points:

```{r simulate-mismatched}
set.seed(42)
n_subj <- 30

# 1. Simulate the GP confounder (Covariate X measured at random times)
tau_cov <- runif(n_subj * 4, 0, 10)
id_cov  <- rep(1:n_subj, each = 4)
# Subject-specific smooth trajectories
X_obs <- sin(tau_cov / 2) + rnorm(n_subj * 4, sd = 0.2)
dat_covariates <- data.frame(id = factor(id_cov), tau = tau_cov, X_obs = X_obs)

# 2. Simulate Treatment Allocation (Measured at a single specific time per subject)
tau_trt <- runif(n_subj, 2, 8)
id_trt  <- 1:n_subj
# True unobserved confounder state at treatment time
X_true_trt <- sin(tau_trt / 2)
# Treatment probability depends on the time-varying confounder
pi_trt <- plogis(-0.5 + 2.0 * X_true_trt)
Trt <- rbinom(n_subj, 1, pi_trt)
dat_treatment <- data.frame(id = factor(id_trt), tau_trt = tau_trt, Trt = Trt, X_gp = 0) # X_gp placeholder

# 3. Simulate Outcome (Measured longitudinally at yet another set of times)
tau_out <- runif(n_subj * 5, 0, 10)
id_out  <- rep(1:n_subj, each = 5)
# True unobserved confounder state at outcome time
X_true_out <- sin(tau_out / 2)
# Outcome confounded by Trt and X
Y <- 2.0 + 1.5 * rep(Trt, each = 5) + 0.8 * X_true_out + rnorm(n_subj * 5, sd = 0.5)
dat_outcome <- data.frame(id = factor(id_out), tau_out = tau_out, Y = Y, X_gp = 0) # X_gp placeholder
```

### Specifying and Fitting the Model

We construct a joint model where `X_obs` informs the `latent_gp`. The GP generates a continuous state `X_gp` which is dynamically queried by the propensity and outcome formulas. 

```{r fit-gp-model}
gp_model <- bjlm_model() |>
  # Treatment block queries X_gp at tau_trt
  propensity(Trt ~ X_gp, data = dat_treatment) |>
  
  # Declare the GP for X_obs across time tau
  latent_gp(name = "X_gp", time = "tau_out", subject = "id", kernel = "se") |>
  
  # Outcome block queries X_gp at tau_out. 
  # We include a random subject intercept (1 | id).
  outcome(
    formula = Y ~ tau_out,
    b0 = ~ 1 + Trt + X_gp + (1 | id),
    b1 = ~ 1,
    data = dat_outcome
  ) |>
  compile()

# Fit using the fast Rust sampler
fit_long <- fit(gp_model, chains = 4, iter = 2000, cores = 4)
```

---

## 3. Post-Processing & Diagnostics

`bjlm` includes a premium suite of post-processing and diagnostic methods tailored to joint longitudinal estimation, allowing you to visualise everything from raw data fits to unobserved confounder trajectories.

### Print and Summary

The `print()` method gives a terminal-friendly overview, while `summary()` provides posterior mean, standard deviation, and R-hat statistics.

```{r summary}
print(fit_long)
summary(fit_long, model = "both")
```

### Predictive Fit

Extract point estimates and credible intervals for the fitted outcome trajectory using `fitted()`.

```{r fitted}
fitted_draws <- fitted(fit_long, summary = TRUE)
head(fitted_draws)
```

### Visualising Model Predictions

Use `plot_predictions()` to overlay posterior fitted trajectories (with 95% credible intervals) on top of the raw data. 

```{r plot-predictions}
# Plot population-level or subject-specific fits
plot_predictions(fit_long, type = "subject", n_subjects = 4)
```

### Visualising Gaussian Processes

The true power of `bjlm` is the continuous-time imputation. You can visualise the inferred, unobserved GP trajectories for individual subjects using `plot_gp()`:

```{r plot-gp}
# Inspect the inferred continuous-time confounder paths
plot_gp(fit_long, n_subjects = 4)
```

### Propensity and Weight Diagnostics

Understand treatment assignment overlap and inverse probability weight distributions using `plot_propensity()`.

```{r plot-propensity}
# Density overlap of propensity scores and distribution of weights
plot_propensity(fit_long, type = "both")
```

---

## 4. Leave-Future-Out Cross Validation

For longitudinal models, standard leave-one-out cross validation often underestimates true predictive error because it attempts to predict past data using future data.

`bjlm` natively supports **Leave-Future-Out Cross Validation (LFO-CV)**. By sequentially predicting time $t+1$ using only data up to time $t$, LFO-CV provides a rigorous assessment of the model's forward-predictive capability. It uses Pareto Smoothed Importance Sampling (PSIS) to approximate the refits, making it incredibly fast.

```{r lfo-cv}
# Perform LFO-CV, starting the forward-validation from time tau = 2.0
lfo_res <- lfo_cv(fit_long, min_tau = 2.0)

# Print the Expected Log Predictive Density (ELPD) summary
print(lfo_res)

# Plot the PSIS Pareto k diagnostic values to assess reliability
plot(lfo_res)
```

High Pareto $k$ values ($k > 0.7$) in the diagnostic plot indicate points where the importance sampling approximation struggles, but `lfo_cv()` will automatically issue warnings when this occurs so you can trust the results.

### Comparing Models with LFO-CV

LFO-CV is especially useful for model selection. Let's fit a competing model that incorrectly omits the treatment effect from the outcome block, and see if LFO-CV correctly identifies the true model (which includes the treatment effect) as having superior predictive performance.

```{r compare-models}
# Fit a competing model omitting the Treatment effect from the outcome
bad_model <- bjlm_model() |>
  propensity(Trt ~ X_gp, data = dat_treatment) |>
  latent_gp(name = "X_gp", time = "tau_out", subject = "id", kernel = "se") |>
  outcome(
    formula = Y ~ tau_out,
    b0 = ~ 1 + X_gp + (1 | id), # Incorrectly omitting Trt
    b1 = ~ 1,
    data = dat_outcome
  ) |>
  compile()

fit_bad <- fit(bad_model, chains = 4, iter = 2000, cores = 4)

# Run LFO-CV on the competing model
lfo_bad <- lfo_cv(fit_bad, min_tau = 2.0)

# Compare ELPD (Expected Log Predictive Density)
cat("True Model ELPD: ", lfo_res$estimates["elpd_lfo", "Estimate"], "\n")
cat("Bad Model ELPD:  ", lfo_bad$estimates["elpd_lfo", "Estimate"], "\n")

# If the 'loo' package is loaded, you can formally compare them:
# loo::loo_compare(list(True_Model = lfo_res, Bad_Model = lfo_bad))
```

As expected, the true model (which correctly specifies the outcome generating process) will yield a significantly higher ELPD score!
