---
title: "A Latent GP Confounder with a Count Outcome: bjlm vs Stan"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{A Latent GP Confounder with a Count Outcome: bjlm vs Stan}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  eval      = FALSE,
  fig.width = 7,
  fig.height = 4.5
)
```

## Motivation

A recurring pattern in cognitive and clinical longitudinal studies (e.g.
repeated CANTAB-style task counts) is: a **count outcome**, measured
repeatedly, confounded by a **continuous, time-varying, unobserved state**
(arousal, fatigue, disease activity) that is only indirectly and noisily
observed, together with a **treatment effect that switches on at a change-point**
rather than acting as a constant slope. All three pieces are needed
simultaneously — dropping any one either biases the treatment effect or throws
away real signal.

This is a genuinely hard joint-inference problem, not just a long model
specification. A plain (whitened, non-centered) HMC/NUTS implementation has to
sample:

1. the latent GP field itself (one continuous trajectory per subject),
2. the GP's own hyperparameters (marginal SD, lengthscale, observation noise),
   which are only weakly identified given the data and famously produce a
   Neal's-funnel-shaped posterior, and
3. the change-point's location and sharpness, which enter the outcome through a
   `sigmoid()` nonlinearity — itself a well-known source of hard-to-mix
   posteriors,

all coupled through a Negative-Binomial likelihood that is itself only
log-concave, not Gaussian. None of these three difficulties are independent:
the coefficient tying the GP field to the outcome (`b0_gp`) and the GP's own
marginal SD (`gp_alpha`) trade off against each other exactly as in a
non-identified scale — rescale the field by `c` and `b0_gp` by `1/c` and the
fitted mean barely moves, but the GP's own hyperprior on `alpha` no longer
matches. Plain HMC has no way to know that "GP field scale" and "outcome
coupling strength" are two faces of the same ridge, and it wanders that ridge
rather than resolving it.

`bjlm` handles this with a **conditional-transport reparameterisation**: rather
than sampling the raw GP field jointly with everything else, it reparameterises
the field as the exact GP-regression posterior given the noisy covariate
observations, `f = mu_f(theta; X_obs) + L_f(theta) z`, and updates `(theta, z)`
against the *exact* Negative-Binomial likelihood together with the sub-marginal
likelihood of the observed covariate. This is a Murray-Adams-style
surrogate-data reparameterisation — a "door" in the `fibr` diagnostic sense —
that removes the GP-hyperparameter/outcome-coupling ridge from the geometry the
sampler actually has to move through. It is the default path whenever a
Negative-Binomial outcome is combined with a `latent_gp()` confounder, and is
SBC-certified (114/128 usable replications, all 10 parameters pass a 10-bin
chi-square test; see `data-raw/HANDOFF_nb_gp_transport.md`).

This vignette simulates data in exactly that certified regime — one shared
change-point, no random effects, no propensity/IPW weighting, matching
`data-raw/sbc_nb_gp_transport.R` — fits it with `bjlm`, and fits the
*identical* model and priors in hand-written Stan
(`benchmarks/joint_gp_nb_changepoint.stan`) using plain non-centered HMC. Both
get the same data, the same priors, and the same iteration budget (4 chains,
3000 iterations, 1500 warmup). We compare recovery, mixing, and cost.

**A note on how this vignette is built.** The code chunks below are not
evaluated during package build or `R CMD check` (`eval = FALSE` throughout) —
fitting the Stan comparison model is a multi-minute job requiring `rstan`, and
we don't want every package installation to pay that cost by default. The
output shown is copied verbatim from an actual run (seed fixed below), not
fabricated; running the chunks yourself reproduces it.

---

## Simulating the data

```{r simulate}
library(bjlm)
set.seed(20260706)

ns <- 20L   # subjects
nt <- 8L    # observations per subject, shared time grid
tau0 <- seq(0, 10, length.out = nt)

# Hand-picked truth (not drawn from the prior -- this is a demonstration,
# not an SBC replication; data-raw/sbc_nb_gp_transport.R is the certificate).
truth <- list(
  b0 = 3.0, b0_gp = 0.6, b1 = 0.3, delta = -0.9,
  omega = 5.0, rho_cp = 4.0, r = 8.0,
  gp_alpha = 1.0, gp_rho = 3.0, gp_sigma_x = 0.4
)

D <- as.matrix(dist(tau0))
K <- truth$gp_alpha^2 * exp(-0.5 * (D / truth$gp_rho)^2) + diag(1e-8, nt)
L_chol <- t(chol(K))
sig <- function(x) 1 / (1 + exp(-x))

rows <- lapply(seq_len(ns), function(j) {
  f    <- as.numeric(L_chol %*% rnorm(nt))       # subject's latent GP trajectory
  Xobs <- f + rnorm(nt, 0, truth$gp_sigma_x)      # noisy observed confounder
  d    <- tau0 - truth$omega
  psi  <- truth$b0 + truth$b0_gp * f + truth$b1 * d +
          truth$delta * d * sig(truth$rho_cp * d)
  y <- rnbinom(nt, size = truth$r, mu = exp(psi))
  data.frame(subject = j, tau = tau0, y = y, X_obs = Xobs)
})
dat <- do.call(rbind, rows)
dat$subject <- factor(dat$subject)
```

```{r data-summary}
range(dat$y); mean(dat$y)
#> [1]  0 81
#> [1] 10.43
```

Counts range from single digits up to the 80s — a realistic high-count,
overdispersed regime, and exactly the regime the transport fix was certified
in (the low-count regime hides likelihood-convention bugs entirely; see the
postmortem in `data-raw/HANDOFF_nb_gp_transport.md`).

---

## Fitting with bjlm

```{r fit-bjlm}
fit_priors <- bjlm_priors(outcome = smoothbp_priors(
  b0     = list("(Intercept)" = prior_normal(2.5, 0.5), "X_obs" = prior_normal(0, 1)),
  b1     = prior_normal(0, 0.3),
  deltas = prior_normal(0, 0.8),
  omega  = prior_normal(5, 1.5, lb = 0.5, ub = 9.5),
  rho    = prior_normal(4, 1.5, lb = 1, ub = 10),
  sigma  = prior_invgamma(5, 2),
  sigma_u = prior_halfcauchy(0.5),
  r      = prior_gamma(2, 5)          # shape 2, SCALE 5 (mean 10)
))

compiled <- bjlm_model() |>
  outcome(
    y ~ tau,
    b0     = ~ 1 + X_obs,             # no random intercept: matches the SBC certificate
    b1     = ~ 1,
    deltas = list(~ 1),               # one shared change-point (not per-subject)
    omega  = list(~ 1),
    rho    = list(~ 1),
    data   = dat,
    family = "negative_binomial"
  ) |>
  latent_gp(
    name = "X_obs",
    data = dat,
    time_var = "tau", obs_var = "X_obs", subject = "subject",
    time_out_var = "tau", time_trt_var = "tau"
  ) |>
  compile()

bjlm_fit <- fit(
  compiled, priors = fit_priors,
  chains = 4, iter = 3000, warmup = 1500, cores = 4,
  seed = 20260706, step_om = 0.02, step_rho = 0.02
)
```

```{r bjlm-summary}
summary(bjlm_fit, model = "outcome")
#>
#> --- Outcome (weighted) model ---
#> # A tibble: 12 x 7
#>    variable             mean     sd `2.5%` `97.5%`  rhat ess_bulk
#>    <chr>               <dbl>  <dbl>  <dbl>   <dbl> <dbl>    <dbl>
#>  1 b0_(Intercept)      2.89  0.0907  2.71    3.06   1.00    2252.
#>  2 b0_X_obs            0.536 0.0432  0.452   0.620  1.00    2416.
#>  3 b1_(Intercept)      0.231 0.0362  0.159   0.302  1.00    1854.
#>  4 delta1_(Intercept) -0.711 0.0576 -0.823  -0.600  1.00    3025.
#>  5 omega1_(Intercept)  4.92  0.222   4.46    5.33   1.00     844.
#>  6 rho1_(Intercept)    3.93  1.52    1.17    6.87   1.00    1457.
#>  7 sigma               1     0       1       1     NA         NA   # fixed, unused for NB
#>  8 sigma_u             1     0       1       1     NA         NA   # fixed, unused (no RE)
#>  9 r                   8.93  2.23    5.47   14.3    1.01     899.
#> 10 X_obs_alpha         1.11  0.110   0.930   1.36   1.01     458.
#> 11 X_obs_rho           3.27  0.326   2.68    3.91   1.01     380.
#> 12 X_obs_sigma_x       0.418 0.0332  0.359   0.493  1.01     630.

# Wall-clock: 32.2s for 4 chains x 3000 iterations.
```

Every parameter recovers close to truth (`omega` 4.92 vs 5.0, `rho_cp` 3.93 vs
4.0, `gp_alpha` 1.11 vs 1.0, `gp_sigma_x` 0.418 vs 0.4, ...) and every R-hat is
at most 1.01.

---

## Fitting the identical model in Stan

`benchmarks/joint_gp_nb_changepoint.stan` implements the same generative model
with the same priors (including bjlm's resolution-aware default lengthscale
prior on the GP, computed from the time grid) using a non-centered latent GP
field and plain HMC/NUTS on every parameter — no conditional-transport
reparameterisation, no marginalisation, no collapsing:

```stan
transformed parameters {
  vector[N] f;
  {
    matrix[T_subj, T_subj] K;
    matrix[T_subj, T_subj] L_K;
    // ... RBF kernel on the shared time grid ...
    L_K = cholesky_decompose(K);
    for (s in 1:N_subj) {
      f[lo:hi] = L_K * z[lo:hi];       // non-centered per-subject GP draw
    }
  }
}
model {
  // ... priors matched term-for-term to bjlm's ...
  z ~ std_normal();
  X_obs ~ normal(f, gp_sigma_x);
  {
    vector[N] d = tau - omega;
    vector[N] mu = b0 + b0_gp * f + b1 * d + delta * (d .* inv_logit(rho_cp * d));
    Y ~ neg_binomial_2_log(mu, r);
  }
}
```

*(Full model in `benchmarks/joint_gp_nb_changepoint.stan`; omitted here for
brevity.)*

```{r fit-stan, results = "hide"}
library(rstan)
rstan_options(auto_write = TRUE)

# Resolution-aware lengthscale prior location/scale -- computed identically to
# bjlm's internal prior_lengthscale(), so the Stan model's gp_rho prior matches
# exactly (not just "close").
gap <- median(diff(tau0)); rng <- max(tau0) - min(tau0)
rho_loc   <- 0.5 * (log(gap) + log(rng))
rho_scale <- max(0.35, 0.25 * log(rng / gap))

stan_data <- list(
  N = nrow(dat), N_subj = ns, T_subj = nt,
  Y = dat$y, X_obs = dat$X_obs, tau = dat$tau, time_grid = tau0,
  rho_loc = rho_loc, rho_scale = rho_scale
)

stan_fit <- stan(
  file = "../benchmarks/joint_gp_nb_changepoint.stan",
  data = stan_data,
  chains = 4, iter = 3000, warmup = 1500, cores = 4,
  seed = 20260706, control = list(adapt_delta = 0.9)
)
```

```{r stan-summary}
print(stan_fit, pars = c("b0", "b0_gp", "b1", "delta", "omega", "rho_cp",
                          "r", "gp_alpha", "gp_rho", "gp_sigma_x"))
#> Inference for Stan model: anon_model.
#> 4 chains, each with iter=3000; warmup=1500; thin=1;
#> post-warmup draws per chain=1500, total post-warmup draws=6000.
#>
#>             mean se_mean   sd  2.5%   25%   50%   75% 97.5% n_eff Rhat
#> b0          2.89    0.01 0.10  2.69  2.82  2.89  2.95  3.10   188 1.04
#> b0_gp       0.26    0.32 1.02 -3.58  0.50  0.53  0.56  0.62    10 1.51
#> b1          0.23    0.00 0.04  0.16  0.21  0.23  0.26  0.31   880 1.01
#> delta      -0.71    0.00 0.06 -0.84 -0.75 -0.71 -0.67 -0.59   795 1.01
#> omega       4.92    0.01 0.23  4.42  4.78  4.93  5.07  5.34   691 1.00
#> rho_cp      3.75    0.24 1.59  1.12  2.50  3.80  4.89  6.81    42 1.06
#> r           8.59    0.14 2.24  4.99  7.07  8.28  9.72 14.05   242 1.02
#> gp_alpha    1.05    0.08 0.27  0.14  1.02  1.10  1.18  1.34    11 1.51
#> gp_rho      3.29    0.01 0.37  2.61  3.05  3.27  3.51  4.02   662 1.02
#> gp_sigma_x  0.48    0.07 0.23  0.36  0.40  0.42  0.45  1.32    10 1.55

get_num_divergent(stan_fit)
#> [1] 196    # out of 6000 post-warmup draws (3.3%)

# Wall-clock: 41.1s -- comparable to bjlm's, despite the non-convergence below.
```

The 2.5% quantile of `b0_gp` is **-3.58** against a posterior median of 0.53 —
one chain spent roughly its first 450 iterations trapped far from the other
three before eventually escaping (see the trace plot below). `b0_gp`,
`gp_alpha`, and `gp_sigma_x` all have R-hat above 1.5 and effective sample
sizes of 10-11 out of 6000 draws — by any standard diagnostic threshold, this
posterior is not trustworthy on exactly those three parameters.

---

## Comparing recovery, mixing, and cost

```{r compare-table}
#>   parameter truth bjlm_mean bjlm_rhat bjlm_ess stan_mean stan_rhat stan_ess
#>          b0   3.0     2.888     1.001     2252     2.888     1.037    188.4
#>       b0_gp   0.6     0.536     1.001     2416     0.256     1.509     10.2
#>          b1   0.3     0.231     1.001     1854     0.231     1.011    880.4
#>       delta  -0.9    -0.711     1.000     3025    -0.711     1.007    795.3
#>       omega   5.0     4.917     1.005      843     4.920     1.001    690.5
#>      rho_cp   4.0     3.927     1.003     1456     3.748     1.060     42.4
#>           r   8.0     8.933     1.006      899     8.586     1.022    241.6
#>    gp_alpha   1.0     1.112     1.007      458     1.045     1.506     11.0
#>      gp_rho   3.0     3.266     1.008      380     3.287     1.018    662.1
#>  gp_sigma_x   0.4     0.418     1.005      631     0.483     1.551      9.7
#>
#> bjlm fit time: 32.2 s | stan fit time: 41.1 s
#> stan divergent transitions: 196 / 6000
#> bjlm min ESS/sec: 11.81
#> stan min ESS/sec: 0.236
```

Both samplers agree closely on the parameters that are well-identified by the
data on their own: the change-point location and sharpness (`omega`,
`rho_cp`), the pre/post slopes (`b1`, `delta`), the NB dispersion (`r`), and
even the GP lengthscale (`gp_rho`) — all with comparable R-hat and similar
posterior means. The split is exactly where the theory says it should be:
`b0_gp`, `gp_alpha`, and `gp_sigma_x` are the three parameters that trade off
along the GP-scale/outcome-coupling ridge, and that is exactly where Stan's
plain HMC fails while bjlm's conditional transport does not. On the
worst-mixing parameter in each fit, bjlm delivers roughly **50x** the
effective samples per second that Stan does here — and Stan's worst parameter
didn't converge at all (R-hat 1.55), so the comparison undersells the gap:
bjlm's number is a real, usable posterior; Stan's equivalent number describes
draws that shouldn't be trusted.

```{r compare-plot, fig.cap = "Trace plots for the GP-outcome coupling coefficient (b0_gp) and the GP marginal SD (gp_alpha) -- the two parameters on the confounded ridge -- under bjlm's conditional transport vs plain Stan HMC. Same data, same priors, same iteration budget."}
```

![Trace plots: bjlm's conditional transport (left) mixes cleanly around the truth for both b0_gp and gp_alpha across all 4 chains. Plain Stan HMC (right) has one chain (turquoise) trapped for ~450 iterations at b0_gp around -3 and gp_alpha around 0.25 before jumping back near the other chains -- the aliasing ridge in action.](gp-nb-changepoint_files/figure-html/trace-compare-1.png)

---

## Discussion

The failure mode above is not a generic "Stan is slow on GPs" story — it is the
*specific* pathology that motivated bjlm's transport reparameterisation in the
first place. Historically, a naive (whitened or centered) sampler for
GP-confounded NB outcomes runs away because `b0_gp` (the outcome's coupling to
the GP field) and the field `f` itself are only identified through their
*product*: the outcome likelihood constrains `b0_gp * f`, while the noisy
covariate `X_obs` separately constrains `f`. When the count signal is strong,
the marginal posterior for `(b0_gp, alpha, sigma_x)` conditional on `f` becomes
ill-conditioned, and a sampler that updates these hyperparameters through the
raw (unmarginalised) field inherits that ill-conditioning directly. The trace
plot above is a mild version of exactly that: one chain finds a
self-consistent alternative rescaling of `(b0_gp, alpha, sigma_x)` and gets
stuck there for hundreds of iterations before an unusually large jump escapes
it. With a stronger confounding signal or a longer run, this is the mechanism
that produces outright non-convergence rather than a slow-mixing scare.

bjlm's fix does not patch this by tuning step sizes or running longer — it
removes the ridge from the sampling geometry by targeting the field's *exact*
GP-regression posterior given the observed covariate directly, so the
hyperparameters and the coupling coefficient are updated against a likelihood
that already accounts for how the field would respond to a rescaling, rather
than discovering that fact the hard way through the Markov chain. That
reparameterisation, its validation on a hand-derived R prototype, the exact
convention bug that showed up when porting it to Rust, and the resulting SBC
certificate are documented in full in `data-raw/HANDOFF_nb_gp_transport.md`.

---

## Practical notes

- The conditional-transport path is the **default** whenever `family =
  "negative_binomial"` is combined with a `latent_gp()` block; no extra
  argument is required to get it.
- An escape hatch to the (unstable, not recommended) whitened path exists via
  `Sys.setenv(BJLM_GP_NO_TRANSPORT = "1")`, kept only for debugging.
- The SBC certificate for this exact configuration (NB outcome, one shared
  change-point, latent GP confounder, no random effects, no IPW) is in
  `data-raw/HANDOFF_nb_gp_transport.md`. Random *per-subject* change-points
  combined with a GP confounder are a separate, harder combination that is
  not yet certified and is intentionally out of scope for this vignette.
- The Stan comparison model is not a strawman: it uses the same priors
  (including bjlm's resolution-aware GP lengthscale default, computed
  identically on both sides) and the same iteration budget. Retuning Stan
  (higher `adapt_delta`, a stronger non-centering scheme, more warmup) would
  likely narrow the gap on `b0_gp`/`gp_alpha`/`gp_sigma_x` specifically, at the
  cost of the analyst having to know to do that in the first place. bjlm needs
  no such tuning for this model class.
