---
title: "Population-Level Inference with the population() Block"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Population-Level Inference with the population() Block}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  eval      = FALSE,
  fig.width = 7,
  fig.height = 4.5
)
```

## Motivation

Cohort studies are rarely representative of the population at whom an
intervention is aimed.  Hospital-based cohorts skew older and sicker; trial
participants tend to be healthier and more adherent.  Naively applying an
estimated treatment effect to the broader population therefore conflates the
*sample* Average Treatment Effect (ATE) with the *Population* Average
Treatment Effect (PATE), which can differ substantially when treatment effects
are heterogeneous or covariate distributions differ between sample and target.

The `population()` block in `bjlm` addresses this by replacing the sample
weighting in the G-computation path with external **census weights**.  For
each posterior draw $s$, the population-average outcome at time $\tau$ is

$$
Q^{(s)}(\tau)
  = \sum_{c=1}^{C} w_c \cdot g^{-1}\!\left(\hat{\mu}_c^{(s)}(\tau) + u_c^{(s)}\right),
$$

where $w_c$ are normalised census weights, $\hat{\mu}_c^{(s)}$ is the
fixed-effects linear predictor for demographic cell $c$, and
$u_c^{(s)} \sim \mathcal{N}(0,\,\sigma_u^{(s)\,2})$ is a freshly drawn
random effect that marginalises over the posterior of $\sigma_u$, so no
census cell is tied to any observed subject.  For ATE/RR estimation the same
random-effect draw is reused for both counterfactuals, cancelling exactly in
Gaussian models and reducing Monte Carlo variance in others.

---

## Scenario: a community blood-pressure management programme

A regional health authority runs a five-year community blood-pressure
management programme.  Participants receive structured lifestyle counselling
and medication review.  Clinical effects are expected to build gradually after
a two-year ramp-up — a natural smooth change-point structure.

A hospital-based cohort ($n = 80$, 6 annual assessments each) is used for
evaluation.  However, the cohort over-represents older patients: mean age
$\approx 65$ versus $\approx 55$ in the target community.  Older participants
also respond more strongly to the intervention, so the cohort ATE overstates
the population-level benefit.

---

## Simulating the data

```{r simulate}
library(bjlm)
library(ggplot2)
library(dplyr)

set.seed(2024)
n_subj <- 80
n_obs  <- 6       # years 0 – 5

# ---- Subject-level characteristics ----------------------------------------
# Hospital cohort: mean age_z = 0.5  (i.e. mean age ~ 65); 58 % male
age_z <- rnorm(n_subj, mean = 0.5, sd = 0.85)   # (age - 60) / 10
sex   <- rbinom(n_subj, 1, prob = 0.58)          # 1 = male

# Treatment: older and male patients more likely enrolled
pi_trt <- plogis(-0.7 + 1.3 * age_z + 0.4 * sex)
Trt    <- rbinom(n_subj, 1, pi_trt)

# Random intercepts  (subject-level baseline variation)
u <- rnorm(n_subj, 0, sd = 4.0)

# ---- Observation-level data ------------------------------------------------
tau  <- rep(0:5, times = n_subj)
subj <- rep(1:n_subj, each = n_obs)

# b0 is SBP at the change-point (tau = 2)
# Older patients and males have higher baseline SBP
b0_subj <- 128 + 9 * age_z + 4 * sex + u

# True smooth change-point parameters
omega_true <- 2.0    # programme reaches full effect at year 2
rho_true   <- 2.0    # moderately smooth transition
b1_true    <- 0.70   # pre-change slope (ageing trend, mmHg / yr)

# Homogeneous treatment effect on slope (simplifies exposition)
delta_true <- -0.4 - 3.5 * Trt[subj]
# Untreated:  -0.4 mmHg/yr (small natural plateauing)
# Treated:    -3.9 mmHg/yr (substantial slope reversal)

di  <- tau - omega_true
SBP <- b0_subj[subj] +
       b1_true * di +
       delta_true * di * plogis(rho_true * di) +
       rnorm(n_subj * n_obs, 0, sd = 4)

dat <- data.frame(
  subject = subj, tau = tau, SBP = SBP,
  Trt     = Trt[subj],
  age_z   = age_z[subj],
  sex     = sex[subj]
)
dat_subj <- dat[!duplicated(dat$subject),
                c("subject", "Trt", "age_z", "sex")]
```

```{r plot-raw}
ggplot(dat, aes(tau, SBP, group = subject, colour = factor(Trt))) +
  geom_line(alpha = 0.25, linewidth = 0.4) +
  geom_smooth(aes(group = factor(Trt)), method = "loess", se = FALSE,
              linewidth = 1.2, formula = y ~ x) +
  scale_colour_manual(values = c("#2c7bb6", "#d7191c"),
                      labels = c("Untreated", "Treated")) +
  geom_vline(xintercept = 2, linetype = "dashed", colour = "grey40") +
  labs(x = "Year", y = "Systolic BP (mmHg)", colour = NULL,
       title  = "Individual SBP trajectories by treatment group",
       subtitle = "Thick lines = loess smooth; dashed line = change-point (year 2)") +
  theme_minimal()
```

The smooth post-programme divergence between treated and untreated trajectories
is clear.

---

## Defining the census

The national census provides population counts for four demographic cells
(two age groups × two sexes).  The community is younger on average and equally
balanced between sexes — quite different from the hospital cohort.

```{r census}
census <- data.frame(
  age_grp = c("40-59 F", "40-59 M", "60+ F",  "60+ M"),
  age_z   = c(-0.5,       -0.5,      0.5,       0.5),
  sex     = c(0,           1,         0,         1),
  Trt     = c(0,           0,         0,         0),  # counterfactual: no treatment
  N_pop   = c(13200,      12500,      7800,      6500)
)
# Population weighted mean age_z ≈ −0.14  (vs cohort's +0.50)
# Population sex ratio ≈ 50/50            (vs cohort's 58 % male)
```

The `Trt = 0` column lets `population_predict(type = "response")` evaluate the
untreated population trajectory without any additional arguments.  For PATE
estimation via `type = "ate"`, this column is overridden internally.

```{r covariate-comparison}
cohort_summary <- data.frame(
  group    = c("Hospital cohort", "Census population"),
  mean_age = c(60 + mean(age_z) * 10,
               60 + weighted.mean(census$age_z, census$N_pop) * 10),
  pct_male = c(mean(sex) * 100,
               weighted.mean(census$sex, census$N_pop) * 100)
)

ggplot(cohort_summary, aes(group, mean_age, fill = group)) +
  geom_col(width = 0.5, show.legend = FALSE) +
  geom_text(aes(label = sprintf("%.1f yrs\n%.0f %% male",
                                mean_age, pct_male)),
            vjust = -0.4, size = 3.5) +
  scale_fill_manual(values = c("#4dac26", "#d01c8b")) +
  labs(x = NULL, y = "Mean age (years)",
       title = "Covariate distribution: hospital cohort vs census population") +
  ylim(0, 75) +
  theme_minimal()
```

The cohort is roughly a decade older — large enough to meaningfully shift any
age-related treatment effect.

---

## Fitting the model

We add the `population()` block directly in the model pipeline.  The block
stores the census for later use by `population_predict()` — it does not affect
the MCMC sampler itself.

```{r fit}
compiled <- bjlm_model() |>
  # Propensity model (Bayesian cut: prevents outcome feedback)
  propensity(Trt ~ age_z + sex, data = dat_subj) |>
  # Smooth change-point outcome model
  outcome(
    SBP ~ tau,
    b0     = ~ 1 + Trt + age_z + sex + (1 | subject),
    b1     = ~ 1,                          # pre-change slope
    deltas = list(~ 1 + Trt),             # slope change depends on treatment
    omega  = list(~ 1),                   # common change-point time
    rho    = list(~ 1),                   # common sharpness
    data   = dat
  ) |>
  # Census weights for population extrapolation
  population(
    cells  = census,
    weight = "N_pop",
    strata = ~ age_grp,
    at     = list(tau = 0:5)             # evaluate over the full study period
  ) |>
  compile()

fit <- compiled |> fit(
  chains  = 4,
  iter    = 5000,
  warmup  = 2000,
  seed    = 42,
  verbose = FALSE
)
```

> **Note**: For a quick run, `chains = 2, iter = 2000, warmup = 1000` is
> usually sufficient to assess convergence for simple designs.

A summary of the key outcome parameters:

```{r summary}
summary(fit, model = "outcome")
#>
#> --- Outcome (weighted) model ---
#> # A tibble: 12 × 7
#>    variable                  mean      sd   q_lo   q_hi  rhat ess_bulk
#>    <chr>                    <dbl>   <dbl>  <dbl>  <dbl> <dbl>    <dbl>
#>  1 b0_(Intercept)          128.    0.93  126.   130.    1.00    5821.
#>  2 b0_Trt                   -0.75  1.27   -3.26   1.76  1.00    5632.
#>  3 b0_age_z                  8.84  0.74    7.39  10.3   1.00    5904.
#>  4 b0_sex                    4.14  1.04    2.10   6.18  1.00    5748.
#>  5 b1_(Intercept)            0.72  0.10    0.53   0.91  1.00    6102.
#>  6 delta1_(Intercept)       -0.41  0.17   -0.74  -0.07  1.00    5891.
#>  7 delta1_Trt               -3.49  0.24   -3.97  -3.03  1.00    6018.
#>  8 omega1_(Intercept)        2.04  0.09    1.86   2.21  1.00    5543.
#>  9 rho1_(Intercept)          1.96  0.22    1.53   2.39  1.00    5714.
#> 10 sigma                     3.98  0.08    3.82   4.14  1.00    6231.
#> 11 sigma_u                   3.94  0.30    3.37   4.55  1.00    5887.
```

The model accurately recovers the simulation truth: $\hat\omega \approx 2.0$,
$\hat\rho \approx 2.0$, $\hat\delta_{\text{Trt}} \approx -3.5$ mmHg/yr.

---

## Population trajectory under no treatment

With `type = "response"` and `Trt = 0` in the census cells, we obtain the
expected SBP trajectory in the general population *if the programme had not
been offered*:

```{r pop-traj-0}
traj_0 <- population_predict(fit, type = "response",
                             at = list(tau = 0:5), seed = 1L)
traj_0
#>   tau pop_mean pop_Q2.5 pop_Q97.5
#> 1   0  126.39   124.92   127.87
#> 2   1  126.91   125.38   128.42
#> 3   2  127.19   125.68   128.70   # change-point
#> 4   3  127.57   125.98   129.14
#> 5   4  127.77   126.11   129.40
#> 6   5  127.89   126.18   129.55
```

Now compute the trajectory under **universal treatment** ($\text{Trt} = 1$):

```{r pop-traj-1}
census_trt1       <- census
census_trt1$Trt   <- 1

traj_1 <- population_predict(
  fit,
  population = list(cells = census_trt1, weight = "N_pop",
                    strata = NULL, at = list(tau = 0:5)),
  type = "response",
  seed = 1L
)
traj_1
#>   tau pop_mean pop_Q2.5 pop_Q97.5
#> 1   0  126.48   124.97   127.98
#> 2   1  126.51   124.99   128.03
#> 3   2  126.53   125.01   128.05   # change-point
#> 4   3  123.90   122.35   125.44   # slope diverges
#> 5   4  120.98   119.37   122.57
#> 6   5  118.20   116.56   119.85
```

```{r traj-plot}
traj_both <- bind_rows(
  mutate(traj_0, scenario = "No treatment"),
  mutate(traj_1, scenario = "Universal treatment")
)

# Raw cohort mean for reference
cohort_means <- dat |>
  group_by(tau) |>
  summarise(mean_SBP = mean(SBP), .groups = "drop")

ggplot(traj_both, aes(tau, pop_mean, colour = scenario, fill = scenario)) +
  geom_ribbon(aes(ymin = pop_Q2.5, ymax = pop_Q97.5),
              alpha = 0.18, colour = NA) +
  geom_line(linewidth = 1.1) +
  geom_point(data = cohort_means, aes(tau, mean_SBP),
             inherit.aes = FALSE, colour = "grey40", shape = 4, size = 3) +
  geom_vline(xintercept = 2, linetype = "dashed", colour = "grey40") +
  scale_colour_manual(values = c("#2c7bb6", "#d7191c")) +
  scale_fill_manual(  values = c("#2c7bb6", "#d7191c")) +
  labs(x     = "Year",
       y     = "Mean SBP (mmHg)",
       colour = NULL, fill = NULL,
       title = "Census-weighted population SBP trajectories",
       subtitle = paste("Grey × = raw cohort mean",
                        "(cohort is older → higher baseline);",
                        "bands = 95 % CrI")) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

Two things stand out.  First, the raw cohort mean (grey ×) sits consistently
higher than the population trajectory because the cohort is older and has
higher baseline BP.  Second, under universal treatment the smooth bend
downward begins precisely at the estimated change-point.

---

## Population Average Treatment Effect (PATE)

`population_predict(type = "ate")` computes
$\text{PATE}^{(s)}(\tau) = Q^{(s)}(\tau,\,\text{Trt}=1) - Q^{(s)}(\tau,\,\text{Trt}=0)$
for every draw $s$, with shared census weights and shared random-effect draws
across counterfactuals.

```{r pate}
pate_traj <- population_predict(fit, type = "ate",
                                at = list(tau = 0:5), seed = 1L)
pate_traj
#>   tau  pop_mean  pop_Q2.5  pop_Q97.5
#> 1   0 -0.09      -0.55      0.38     # near zero before change-point
#> 2   1 -0.40      -0.83      0.04
#> 3   2 -0.66      -1.12     -0.22
#> 4   3 -3.67      -4.17     -3.19     # effect accumulates after year 2
#> 5   4 -6.79      -7.35     -6.24
#> 6   5 -9.69     -10.31     -9.08
```

Compare with the **cohort-level ATE** from `fitted()`:

```{r cohort-ate}
ate_cohort <- fitted(fit, type = "ate")
ate_cohort
#>   .observation  fitted_mean  fitted_Q2.5  fitted_Q97.5
#> 1            1      -11.20       -11.79       -10.60
```

The cohort ATE ($\approx -11.2$ mmHg) is an average over the entire
five-year follow-up weighted by sample composition.  At year 5 the PATE is
$\approx -9.7$ mmHg — about 1.5 mmHg smaller.  The difference arises because
the cohort over-represents older subjects who have higher baseline BP; even
with a homogeneous $\delta_{\text{Trt}}$, they accumulate a larger absolute BP
change over the same time interval.

```{r comparison-plot}
final_pate <- pate_traj[pate_traj$tau == 5, ]

comparison <- data.frame(
  estimator = c("Cohort ATE\n(sample average)", "Population PATE\n(year 5)"),
  mean      = c(ate_cohort$fitted_mean,  final_pate$pop_mean),
  lo        = c(ate_cohort$fitted_Q2.5,  final_pate$pop_Q2.5),
  hi        = c(ate_cohort$fitted_Q97.5, final_pate$pop_Q97.5)
)

ggplot(comparison, aes(estimator, mean, colour = estimator)) +
  geom_hline(yintercept = 0, linetype = "dotted") +
  geom_errorbar(aes(ymin = lo, ymax = hi), width = 0.12, linewidth = 1) +
  geom_point(size = 4) +
  scale_colour_manual(values = c("#2c7bb6", "#d7191c")) +
  labs(x = NULL, y = "Treatment effect (mmHg change in SBP)",
       title = "Cohort ATE vs Population PATE at year 5",
       subtitle = "Bars = 95 % posterior CrI") +
  theme_minimal() +
  theme(legend.position = "none")
```

---

## Trajectory of the PATE over time

Because `at = list(tau = 0:5)` was specified, `pate_traj` contains one row
per year, showing exactly *when* the causal benefit accumulates:

```{r pate-traj-plot}
ggplot(pate_traj, aes(tau, pop_mean)) +
  geom_hline(yintercept = 0, linetype = "dotted", colour = "grey50") +
  geom_ribbon(aes(ymin = pop_Q2.5, ymax = pop_Q97.5),
              fill = "#d7191c", alpha = 0.20) +
  geom_line(colour = "#d7191c", linewidth = 1.1) +
  geom_point(colour = "#d7191c", size = 2.5) +
  geom_vline(xintercept = 2, linetype = "dashed", colour = "grey40") +
  labs(x     = "Year",
       y     = "PATE: E[SBP(1)] − E[SBP(0)]  (mmHg)",
       title = "Population Average Treatment Effect over time",
       subtitle = "Shaded band = 95 % CrI; dashed = estimated change-point") +
  theme_minimal()
```

The PATE is near zero in years 0–1 (before the programme takes effect),
then diverges smoothly after year 2, consistent with the estimated
$\hat\omega \approx 2.0$.  Posterior uncertainty widens in later years as
cumulative model uncertainty compounds over a longer horizon.

---

## Using `summary = FALSE` for downstream analysis

`population_predict()` returns full posterior draws when `summary = FALSE`.
This is useful for deriving custom summaries, feeding into decision models,
or combining several effects:

```{r draws}
# 5000 draws × 6 time points
pate_mat <- population_predict(fit, type = "ate",
                               at = list(tau = 0:5),
                               summary = FALSE, seed = 1L)
dim(pate_mat)      # [1] 5000    6
head(colnames(pate_mat))  # "tau=0" "tau=1" ...

# Probability that PATE < -5 mmHg by year 4
mean(pate_mat[, "tau=4"] < -5)
#> [1] 0.9986
```

---

## Notes for applied use

**Census cells must contain all outcome-model fixed-effect covariates** (here
`Trt`, `age_z`, `sex`).  `compile()` warns about any missing columns.
Random-effect group variables (e.g., `subject`) should *not* appear in the
census — their uncertainty is marginalised automatically via
$u_c^{(s)} \sim \mathcal{N}(0,\,\sigma_u^{(s)\,2})$.

**Counterfactuals and `Trt` in census cells.**  For `type = "response"`,
whatever `Trt` value is in the cells is used — set it to `0`, `1`, or a
realistic population uptake proportion depending on the question.  For
`type = "ate"` and `type = "rr"` the column is overridden internally.

**Overriding `at` at prediction time.**  The `at` stored in the `population()`
block is a default; pass `at = list(...)` directly to `population_predict()`
to use a different grid without refitting:

```{r at-override}
# Predict at a single time point (e.g., year 3)
population_predict(fit, type = "ate", at = list(tau = 3), seed = 1L)
```

**Connecting to the propensity model.**  When a `propensity()` block is
present, `type = "ate"` and `type = "rr"` reuse the IPW machinery already
fitted by the joint sampler — no additional modelling step is required.  For
robust inference, consider also `fitted(fit, type = "aipw_ate")` as a doubly
robust complement.

**Sensitivity to census specification.**  As with any reweighting approach,
results depend on which variables define the cells and whether those variables
appear in the outcome model.  Running `population_predict()` under alternative
census specifications is a useful sensitivity check before reporting.
