---
title: "Breakpoint Selection with Spike-and-Slab Priors"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Breakpoint Selection with Spike-and-Slab Priors}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  eval      = FALSE,
  fig.width = 7,
  fig.height = 4.5
)
```

## Motivation

Piecewise regression models invite a natural question: *which coefficients in the
change-point component are actually non-zero?*  In a confounded longitudinal
study we might ask not only "does this treatment change the trajectory?" but
also "which baseline characteristics modify the magnitude of that change?"
Including irrelevant modifiers dilutes the precision of everything else;
excluding real ones biases the causal estimate.

`bjlm` resolves this with a **spike-and-slab prior** (Kuo & Mallick 1998) on
the $b_1$ and $\delta_k$ coefficients.  A binary indicator $\gamma_j \in
\{0, 1\}$ gates each coefficient:

$$
\beta_j \mid \gamma_j = \begin{cases}
  0 & \gamma_j = 0 \text{ (spike)}\\
  \mathcal{N}(0,\, \sigma_\text{slab}^2) & \gamma_j = 1 \text{ (slab)}
\end{cases}
$$

The posterior mean of $\gamma_j$ across all MCMC draws is the **Posterior
Inclusion Probability** (PIP): values near 1 indicate strong evidence for
inclusion, near 0 for exclusion, near 0.5 for genuine uncertainty.

Crucially, the gamma samplers operate *within* the IPW-weighted outcome
block, so variable selection is performed conditional on the propensity-score
model.  The Bayesian Cut is preserved — the spike-and-slab step never feeds
back into propensity estimation.

The Rust implementation (`SpikeSlabConfig`, `sample_pi`, `sample_gamma`,
`run_chain_ss`) and the `smoothbp_spike_slab` / `smoothbp_pip` R classes are
adapted from the **smoothbp** package (Bindoff 2025).

---

## Scenario: rehabilitation after hip fracture

Sixty patients are followed for 12 months after hip fracture repair.
Functional mobility (outcome $Y$, higher = better) is assessed at irregular
intervals.  A structured physiotherapy programme (treatment `Trt`) is offered
but uptake is confounded by pre-operative baseline severity (`sev`): frailer
patients are less likely to receive treatment.

We expect recovery to follow a smooth change-point shape — rapid early gains
then plateau — with the change occurring around month 6.  The research
question is: **which covariates genuinely modify the slope change at the
change-point?**

We build the `delta` formula with four candidate modifiers: `Trt` (treatment),
`age_z` (standardised age), `male` (sex), and `sev` (baseline severity).  The
**true** model has non-zero effects for `Trt` and `sev`; `age_z` and `male`
are null.

```{r simulate}
library(bjlm)
library(ggplot2)

set.seed(2025)

n_subj <- 60
n_obs  <- 6  # observations per subject (irregular timing)

# Subject-level characteristics
subj <- data.frame(
  id    = factor(1:n_subj),
  sev   = rnorm(n_subj),            # pre-operative severity (confounder)
  age_z = rnorm(n_subj),            # standardised age (null modifier)
  male  = rbinom(n_subj, 1, 0.45)   # sex (null modifier)
)

# Treatment assignment: confounded by severity
subj$pi_trt <- plogis(-0.4 + 1.2 * subj$sev)
subj$Trt    <- rbinom(n_subj, 1, subj$pi_trt)

# True outcome parameters
true_b0     <-  20     # baseline intercept
true_b0_trt <-   2     # treatment shift in level
true_b0_sev <-  -3     # severity depresses baseline
true_b1     <-   3.0   # initial slope (improvement per month)
true_omega  <-   6     # change-point location
true_rho    <-   1.2   # transition sharpness
true_delta_int <- -2.5 # slope change at omega (all subjects)
true_delta_trt <-  2.0 # treatment modifies the slope change   *** real ***
true_delta_sev <- -1.0 # severity modifies slope change        *** real ***
# true_delta_age  = 0  # age has NO effect on slope change
# true_delta_male = 0  # sex has NO effect on slope change
true_sigma_u <- 2.5    # subject random-effect SD
true_sigma   <- 1.2    # residual SD

# Generate longitudinal observations
dat <- do.call(rbind, lapply(1:n_subj, function(i) {
  tau <- sort(runif(n_obs, 0.5, 12))
  u_i <- rnorm(1, 0, true_sigma_u)
  d   <- tau - true_omega
  s   <- plogis(d * true_rho)
  mu  <- true_b0 + u_i +
         true_b0_trt * subj$Trt[i] +
         true_b0_sev * subj$sev[i] +
         true_b1 * d +
         (true_delta_int +
          true_delta_trt * subj$Trt[i] +
          true_delta_sev * subj$sev[i]) * d * s
  Y <- rnorm(n_obs, mu, true_sigma)
  data.frame(
    id    = subj$id[i],
    tau   = tau,
    Y     = Y,
    Trt   = subj$Trt[i],
    sev   = subj$sev[i],
    age_z = subj$age_z[i],
    male  = subj$male[i]
  )
}))

head(dat)
```

```{r plot-data}
ggplot(dat, aes(tau, Y, group = id, colour = factor(Trt))) +
  geom_line(alpha = 0.4, linewidth = 0.4) +
  geom_point(size = 0.8, alpha = 0.6) +
  scale_colour_manual(values = c("#E69F00", "#56B4E9"),
                      labels = c("Control", "Treated")) +
  labs(x = "Months post-surgery", y = "Functional mobility score",
       colour = "Treatment", title = "Individual recovery trajectories") +
  theme_minimal()
```

The treated trajectories (blue) show a stronger and more sustained plateau
after the change-point — consistent with the true `delta_Trt > 0`.

---

## Fitting the standard model (no variable selection)

Before introducing spike-and-slab, we fit the full model with all four
candidate modifiers.  This is the reference against which we compare
posteriors.

```{r fit-standard}
compiled_model <- bjlm_model() |>
  propensity(Trt ~ sev, data = dat) |>
  outcome(
    Y ~ tau,
    b0     = ~ 1 + Trt + sev + (1 | id),
    b1     = ~ 1,
    deltas = list(~ 1 + Trt + sev + age_z + male),
    omega  = list(~ 1),
    rho    = list(~ 1),
    data   = dat
  ) |>
  compile()

fit_full <- fit(compiled_model, chains = 4, iter = 2000)
```

```{r summary-standard}
summary(fit_full)
```

The full-model posteriors for `delta1_age_z` and `delta1_male` should be
centred near zero with wide credible intervals — evidence for their
irrelevance, but not a firm decision.

---

## Fitting the spike-and-slab model

Passing `spike = prior_spike_slab(...)` to `fit()` adds the
Kuo-Mallick gamma sampler to the outcome block.  The slab prior is a
$\mathcal{N}(0, 2)$ by default; here we also ask the model to learn the
global inclusion probability $\pi$ from the data.

```{r fit-ss}
fit_ss <- fit(
  compiled_model,
  chains = 4,
  iter   = 2000,
  spike  = prior_spike_slab(
    pi       = 0.5,       # initial inclusion probability
    slab     = prior_normal(0, 2),
    learn_pi = TRUE,      # place Beta(1,1) hyperprior on pi
    a = 1, b = 1
  )
)
```

The gamma indicators ($\gamma_j \in \{0,1\}$) for each eligible coefficient
are stored alongside the standard outcome parameters.  Coefficient draws that
belong to iterations where $\gamma_j = 0$ contribute zero to the predicted
mean, so the sampler automatically down-weights irrelevant paths.

---

## Posterior Inclusion Probabilities

```{r pip}
p <- pip(fit_ss)
p
```

```{r plot-pip}
plot(p) +
  labs(title = "Posterior Inclusion Probabilities",
       subtitle = "Grey dashed line = 0.5 decision threshold")
```

The PIP table and plot show:

| Parameter | Expected | Interpretation |
|-----------|----------|----------------|
| `b1_(Intercept)` | high | There is an initial slope |
| `delta1_(Intercept)` | high | The change-point affects all subjects |
| `delta1_Trt` | high | Treatment genuinely modifies the slope change |
| `delta1_sev` | high | Severity genuinely modifies the slope change |
| `delta1_age_z` | low | No evidence that age modifies the change-point |
| `delta1_male` | low | No evidence that sex modifies the change-point |

A PIP > 0.5 is typically taken as evidence for inclusion; values > 0.75 or
< 0.25 represent reasonably strong evidence in either direction.  The 95 %
credible intervals on each PIP (derived from a Beta posterior) are shown as
horizontal error bars.

---

## Comparing posteriors: full model vs. spike-and-slab

The spike-and-slab model does not simply drop variables — it marginalises
over their inclusion, which regularises the posteriors for the retained
coefficients.

```{r compare-posteriors}
library(posterior)

# Extract draws for the real change-point modifier
draws_full <- subset_draws(fit_full$draws,  variable = "delta1_Trt")
draws_ss   <- subset_draws(fit_ss$draws,    variable = "delta1_Trt")

# Overlay posteriors
df_compare <- rbind(
  data.frame(model = "Full model", delta_Trt = as.numeric(draws_full)),
  data.frame(model = "Spike-and-slab", delta_Trt = as.numeric(draws_ss))
)

ggplot(df_compare, aes(delta_Trt, colour = model, fill = model)) +
  geom_density(alpha = 0.25, linewidth = 0.8) +
  geom_vline(xintercept = true_delta_trt, linetype = "dashed") +
  annotate("text", x = true_delta_trt + 0.1, y = 0, label = "truth",
           hjust = 0, vjust = -0.5, size = 3) +
  labs(x = expression(delta["1,Trt"]),
       y = "Density",
       title = "Posterior for treatment slope-change modifier",
       colour = NULL, fill = NULL) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

Because `delta1_age_z` and `delta1_male` are regularised toward zero, the
spike-and-slab posterior for `delta1_Trt` is typically narrower than the
full-model posterior — the model no longer has to share uncertainty with
irrelevant modifiers.

---

## Causal effects under spike-and-slab

The `fitted()` and `causal_effect()` post-processing functions work
unchanged on a spike-and-slab fit.  The gamma indicators are marginalised
out in the prediction step: draws from zero-gated iterations contribute zero
slope-change, so the causal-effect posterior naturally incorporates
model-selection uncertainty.

```{r causal-effect}
# G-computation ATE averaged over the posterior
fitted(fit_ss, type = "ate")

# Doubly robust AIPW ATE
fitted(fit_ss, type = "aipw_ate")
```

---

## Sensitivity: fixed vs. learned pi

When `learn_pi = FALSE` (the default), $\pi$ is fixed at the value passed
to `prior_spike_slab()`.  When `learn_pi = TRUE`, $\pi$ receives a
`Beta(a, b)` hyperprior and is updated each iteration.  Comparing PIPs under
both settings is a useful sensitivity check:

```{r sensitivity-pi}
fit_fixed_pi <- fit(
  compiled_model,
  chains = 4, iter = 2000,
  spike  = prior_spike_slab(pi = 0.5, learn_pi = FALSE)
)

fit_learn_pi <- fit(
  compiled_model,
  chains = 4, iter = 2000,
  spike  = prior_spike_slab(pi = 0.5, learn_pi = TRUE, a = 1, b = 1)
)

pip_fixed <- pip(fit_fixed_pi)
pip_learn <- pip(fit_learn_pi)

pip_fixed$pi_setting <- "Fixed (pi = 0.5)"
pip_learn$pi_setting <- "Learned (Beta(1,1))"

pip_all <- rbind(pip_fixed, pip_learn)

ggplot(pip_all, aes(pip, reorder(parameter, pip),
                    colour = pi_setting, shape = pi_setting)) +
  geom_vline(xintercept = 0.5, linetype = "dashed", alpha = 0.4) +
  geom_point(size = 3, position = ggplot2::position_dodge(width = 0.4)) +
  scale_x_continuous(limits = c(0, 1)) +
  labs(x = "PIP", y = NULL, colour = NULL, shape = NULL,
       title = "PIP sensitivity to pi hyperprior") +
  theme_minimal() +
  theme(legend.position = "bottom")
```

Robust conclusions — large PIPs near 1 or small PIPs near 0 — are stable
across choices of $\pi$.  PIPs near 0.5 in both settings indicate genuine
uncertainty that additional data would be needed to resolve.

---

## Summary

| Step | Code |
|------|------|
| Declare spike prior | `prior_spike_slab(pi = 0.5, learn_pi = TRUE)` |
| Fit with selection | `fit(compiled_model, spike = prior_spike_slab(...))` |
| Extract PIPs | `pip(fit_ss)` |
| Visualise PIPs | `plot(pip(fit_ss))` |
| Causal effects | `fitted(fit_ss, type = "aipw_ate")` |

The spike-and-slab path in `bjlm` adds one argument to `fit()` and one
post-processing call.  The sampler, the propensity model, the Bayesian Cut,
and all causal-effect estimators are otherwise unchanged.
